---
title: "UK GDP smoothing"
author: "Adri Pérez"
date: "4/12/2024"
output: html_document

astsa library version: install.packages("https://cran.r-project.org/src/contrib/Archive/astsa/astsa_1.12.tar.gz", repos = NULL, type = "source")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Import relevant datasets.

GDP \<- read_excel("C:.../GDP24.xls",

sheet = "SAS", col_types = c("date", "numeric"))

#GDP\$Date \<- as.Date(as.yearqtr(GDP\$Date, format = '%Y Q%q')) -1

GDP\$Date \<- as.Date(GDP\$Date)

```{r}
library(readxl)
library(zoo)
library(astsa)
library(ggplot2)
library(writexl)
library(dplyr)


# status: updated
GDP <- read_excel("C:.../GDP.xls", 
     sheet = "SAS", col_types = c("date", "numeric"))
#GDP$Date <- as.Date(as.yearqtr(GDP$Date, format = '%Y Q%q')) -1
GDP$Date <- as.Date(GDP$Date)

# satus: updated
UR <- read_excel("C:.../UR.xls", 
     sheet = "SAS", col_types = c("date", "numeric"))
UR$Date <- as.Date(UR$Date)

# satus: updated
MACRO_IND <- read_excel("C:.../datos UK.xlsX", 
     sheet = "SAS", col_types = c("date", "numeric", "numeric", "numeric", "numeric"))
MACRO_IND$Dates <- as.Date(MACRO_IND$Dates) 

# satus: updated
CPIH <- read_excel("C:.../CPIH.xls", 
     sheet = "SAS", col_types = c("text", "numeric"))
CPIH$Date <- as.Date(as.yearqtr(CPIH$Date, format = '%Y Q%q')) -1

# satus: updated
GDI <- read_excel("C:.../Deuda y RBD hogares.xlsx", 
     sheet = "SAS", col_types = c("date", "numeric", "numeric"))
GDI$Date <- as.Date(GDI$Date)
```

Clean Macro Indicadors data.

\- Rename date column

\- Transform all quarterly dates to last day of month to avoid missings when merging.

```{r}
library(lubridate)
# rename column
names(MACRO_IND)[names(MACRO_IND) == "Dates"] <- "Date"
# transform date
MACRO_IND$Date <- ceiling_date(MACRO_IND$Date, "month") - days(1)
```

Join all data into a single dataset.

```{r}
# rename macro indicators date column
options(max.print = 1e6) 

df <- merge(x=GDP,y=UR, by="Date", all.x=TRUE)
df <- merge(x=df,y=MACRO_IND, by="Date", all.x=TRUE)
df <- merge(x=df,y=GDI, by="Date", all.x=TRUE)
df <- merge(x=df,y=CPIH, by="Date", all.x=TRUE)
df <- df[order(df$Date), ]
write_xlsx(df, "C:.../df_baseline_new.xlsx")
```

Compute the annual rates for GDP, employed, VAB and business income for its use in the state space models.

```{r}

df["GDP quarterly growth rate"] <- c(NA, diff(log(df$GDP),1)) * 100
df["Employed quarterly growth rate"] <- c(NA, diff(log(df$Employed),1)) * 100
df["Business Income quarterly growth rate"] <- c(NA, diff(log(df$`Business income`),1)) * 100
df["Gross Disposable Income quarterly growth rate"] <- c(NA, diff(log(df$GDI),1)) * 100
df["Consumer Price Index quarterly growth rate"] <- c(NA, diff(log(df$CPIH),1)) * 100
df["UR differences"] <- c(NA, diff(df$UR,1)) + 0.000001
df["Consumer_confi_diff"] <- c(NA, diff(df$`Consumer confidence`,1))
```

Check the correlation structure between the variables of interest.

Including COVID data in the GDP annual growth rate:

```{r}

corr_matrix <- as.matrix(cor(df[c("GDP quarterly growth rate", "UR differences", "Employed quarterly growth rate", "Business Income quarterly growth rate", "Consumer confidence", "Gross Disposable Income quarterly growth rate", "Consumer Price Index quarterly growth rate")], use="pairwise.complete.obs"))

corr_df <- as.data.frame(corr_matrix)

write_xlsx(corr_df, "C:.../PD_series/outputs/corr_macro_new.xlsx")
```

Excluding COVID data from GDP. It is considered that the COVID effect goes from March 20 (pandemic breakthrough) to March 22.

```{r}
df["GDP quarterly growth rate 2"] <- ifelse(df$Date >= "2020-03-31" & df$Date <= "2022-03-31", NA, df$`GDP quarterly growth rate`)

corr_matrix <- as.matrix(cor(df[c("GDP quarterly growth rate 2", "UR differences" ,"Employed quarterly growth rate", "Business Income quarterly growth rate", "Consumer_confi_diff", "Gross Disposable Income quarterly growth rate", "Consumer Price Index quarterly growth rate")], use="pairwise.complete.obs"))

corr_df <- as.data.frame(corr_matrix)

write_xlsx(corr_df, "C:.../PD_series/outputs/corr_excl_covid_new.xlsx")
```

It is decided to exclude the UR from the state space model given that it has a small correlation with the GDP rate and, in addition, it is against the economic theory.

```{r}
#Data Preparation
series_0=subset(df[c("Date", "GDP quarterly growth rate 2", "UR differences" ,"Employed quarterly growth rate", "Business Income quarterly growth rate")], df$Date >= "1980-09-30")

write_xlsx(series_0, "C:.../PD_series/outputs/series0_new.xlsx")

series <- series_0[,-1]

N=nrow(series)
p=ncol(series)
y=as.matrix(series,nc=p)
y[is.na(y)]<- 0

A = array(0, dim=c(p,p,N))
for(k in 1:N) {
  if (y[k,1] != 0) A[1,1,k]= 1
  if (y[k,2] != 0) A[2,2,k]= 1
  if (y[k,3] != 0) A[3,3,k]= 1
  if (y[k,4] != 0) A[4,4,k]= 1
  }

mu0 = matrix(y[1,], nr=p, nc=1)
Sigma0 = diag(c(0.005, 0.005), p)

Phi = diag(1, p)
cQ = diag(c(0.5, 0.5), p)
cR = diag(c(0.05, 0.05), p)

library(astsa)

em = EM1(N, y, A, mu0, Sigma0, Phi, cQ, cR, 7000, .00001)
```

```{r}
(mu0=em$mu0)
(Sigma0=em$Sigma0)
(Phi=em$Phi)
(cQ=chol(em$Q))
(cR=chol(em$R))
```

```{r}
y.kalman<-Ksmooth1(N,y,A,mu0,Sigma0,Phi,0,0,cQ,cR,0)

p1 = 2*sqrt(y.kalman$Ps[1,1,])
p2 = 2*sqrt(y.kalman$Ps[2,2,])
p3 = 2*sqrt(y.kalman$Ps[3,3,])
p4 = 2*sqrt(y.kalman$Ps[4,4,])

pf1 = 2*sqrt(y.kalman$Pf[1,1,])
pf2 = 2*sqrt(y.kalman$Pf[2,2,])
pf3 = 2*sqrt(y.kalman$Pf[3,3,])
pf4 = 2*sqrt(y.kalman$Pf[4,4,])
```

```{r}
smoothings <- ts(matrix(c(y.kalman$xs),nc=p,byrow = T),start=c(1980,3),freq=4)
original <- ts(series[,1],start=c(1980,3),freq=4)
plot(smoothings[,1],col=1, ylim=c(-2.5,2.5))
lines(smoothings[,1]+p1, lty=2, col=4)
lines(smoothings[,1]-p1, lty=2, col=4)
lines(original, lty=3, col=2)

time_index <- time(smoothings)

# Convert smoothings to a data frame
smoothings_df <- as.data.frame(smoothings)
smoothings_df$Date <- as.yearqtr(time_index)  
smoothings_df <- smoothings_df[, c(ncol(smoothings_df), 1:(ncol(smoothings_df)-1))]

write_xlsx(smoothings_df, "C:.../PD_series/outputs/smoothings_new.xlsx")


```

```{r}
filterings <- ts(matrix(c(y.kalman$xf),nc=p,byrow = T),start=c(1980,3),freq=4)
plot(filterings[,1],col=1, ylim=c(-2.5,2))
lines(filterings[,1]+pf1, lty=2, col=4)
lines(filterings[,1]-pf1, lty=2, col=4)
lines(original, lty=3, col=2)

time_index <- time(filterings)

# Convert smoothings to a data frame
filterings_df <- as.data.frame(filterings)
filterings_df$Date <- as.yearqtr(time_index)  
filterings_df <- filterings_df[, c(ncol(filterings_df), 1:(ncol(filterings_df)-1))]

write_xlsx(filterings_df, "C:.../PD_series/outputs/filterings_new.xlsx")
```

```{r}
plotting_ds <- subset(df, df$Date >= "1980-09-30")[, c("Date", "GDP quarterly growth rate")]
plotting_ds[,"GDP quarterly growth rate smoothed"] <- smoothings[,1]
plotting_ds[,"GDP quarterly growth rate filtered"] <- filterings[,1]


p <- ggplot(plotting_ds) +
  geom_line( color="#000000", aes(x=Date, y=`GDP quarterly growth rate`), ) + 
  geom_line( color="#0096C8", aes(x=Date, y=`GDP quarterly growth rate smoothed`)) +   
  geom_line( color="#FF8728", aes(x=Date, y=`GDP quarterly growth rate filtered`)) + 
  xlab("Date") +
  theme_light()

p
```

```{r}
plotting_ds["synthetic GDP"] = ifelse(plotting_ds$Date >= "2020-03-31" & plotting_ds$Date <= "2022-03-31", plotting_ds$`GDP quarterly growth rate smoothed`, plotting_ds$`GDP quarterly growth rate`)

p <- ggplot(plotting_ds) +
  geom_rect(aes(xmin = as.Date("2019-12-31"), xmax = as.Date("2022-03-31"),
                ymin = -Inf, ymax = Inf), fill = "lightblue", alpha = 0.05) +
  geom_area(aes(x = Date, 
                y = ifelse(`synthetic GDP` < 0, `synthetic GDP`, 0)), 
            fill = "red", alpha = 0.5) +  
  geom_line( color="#006DFF", aes(x=Date, y=`synthetic GDP`)) +   
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1) + 
  geom_text(aes(x = as.Date("2021-03-01"), y = 1.5, label = "COVID"), 
            color = "Black", size = 4, angle = 0, fontface="italic") +  
  xlab("Date") +
  ylab("GDP (quarterly growth rate)") +
  theme_light()

p
```

```{r}
ggsave(filename = "GDP synthetic (UR & Employed & Business Income).png", plot = p, width = 8, height = 6, dpi = 300)
```

```{r}
plotting_ds["synthetic GDP 2"] = ifelse(plotting_ds$Date >= "2020-03-31" & plotting_ds$Date <= "2022-03-31", plotting_ds$`GDP quarterly growth rate filtered`, plotting_ds$`GDP quarterly growth rate`)

p <- ggplot(plotting_ds) +
  geom_rect(aes(xmin = as.Date("2019-12-31"), xmax = as.Date("2022-03-31"),
                ymin = -Inf, ymax = Inf), fill = "lightblue", alpha = 0.05) +
  geom_area(aes(x = Date, 
                y = ifelse(`synthetic GDP 2` < 0, `synthetic GDP 2`, 0)), 
            fill = "red", alpha = 0.5) +  
  geom_line( color="#006DFF", aes(x=Date, y=`synthetic GDP 2`)) +   
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1) + 
  geom_text(aes(x = as.Date("2021-03-01"), y = 1.5, label = "COVID"), 
            color = "Black", size = 4, angle = 0, fontface="italic") +  
  xlab("Date") +
  ylab("GDP (quarterly growth rate)") +
  theme_light()
  2
p

```

```{r}
# this table goes to the report
cor(smoothings)
```

```{r}
cor(filterings)
```

```{r}
cor(series, use = "pairwise.complete.obs")
```

```{r}
series
```

```{r}
plotting_ds["GDP annual growth rate"] <- (c(NA, NA, NA, rollapply(plotting_ds["synthetic GDP"]/100 + 1, width = 4, FUN = prod, align = "right") - 1))*100

```

```{r}
p <- ggplot(subset(plotting_ds, plotting_ds$Date >= "1981-06-30")) +
  geom_rect(aes(xmin = as.Date("2019-12-31"), xmax = as.Date("2022-03-31"),
                ymin = -Inf, ymax = Inf), fill = "lightblue", alpha = 0.05) +
  geom_area(aes(x = Date, 
                y = ifelse(`GDP annual growth rate` < 0, `GDP annual growth rate`, 0)), 
            fill = "red", alpha = 0.5) +  
  geom_line( color="#006DFF", aes(x=Date, y=`GDP annual growth rate`)) +   
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1) + 
  geom_text(aes(x = as.Date("2021-03-01"), y = 4, label = "COVID"), 
            color = "Black", size = 4, angle = 0, fontface="italic") +  
  xlab("Date") +
  ylab("GDP (annual growth rate)") +
  theme_light()

p
```

```{r}
ggsave(filename = "Annual GDP synthetic (UR & Employed & Business Income).png", plot = p, width = 8, height = 6, dpi = 300)
```

```{r}
g <- ts(plotting_ds$`synthetic GDP`,start=c(1980,3),freq=4)/100+1
g1 <- lag(g,-1)
g2 <- lag(g,-2)
g3 <- lag(g,-3)
g4 <- lag(g,-4)
g5 <- lag(g,-5)
g6 <- lag(g,-6)
```

```{r}
gdp_pa <- ((1+(g2)+(g2*g1)+(g2*g1*g))/((1/(g3))+(1/(g3*g4))+(1/(g3*g4*g5))+(1/(g3*g4*g5*g6)))-1)*100
plotting_ds["gdp_pa"] <- c(NA, NA, NA, NA, NA, NA, gdp_pa)
```

```{r}
# add gdp pa with original GDP
g <- ts(plotting_ds$`GDP quarterly growth rate`,start=c(1980,3),freq=4)/100+1
g1 <- lag(g,-1)
g2 <- lag(g,-2)
g3 <- lag(g,-3)
g4 <- lag(g,-4)
g5 <- lag(g,-5)
g6 <- lag(g,-6)

gdp_pa_0 <- ((1+(g2)+(g2*g1)+(g2*g1*g))/((1/(g3))+(1/(g3*g4))+(1/(g3*g4*g5))+(1/(g3*g4*g5*g6)))-1)*100
plotting_ds["gdp_pa_0"] <- c(NA, NA, NA, NA, NA, NA, gdp_pa_0)

```

```{r}
p <- ggplot(subset(plotting_ds, plotting_ds$Date >= "1982-03-31")) +
  geom_rect(aes(xmin = as.Date("2019-12-31"), xmax = as.Date("2022-03-31"),
                ymin = -Inf, ymax = Inf), fill = "lightblue", alpha = 0.05) +
  geom_area(aes(x = Date, 
                y = ifelse(`gdp_pa` < 0, `gdp_pa`, 0)), 
            fill = "red", alpha = 0.5) +  
  geom_line( color="#006DFF", aes(x=Date, y=`gdp_pa`)) + 
  geom_line(color = "black", aes(x = Date, y = `gdp_pa_0`)) +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed", size = 1) + 
  geom_text(aes(x = as.Date("2019-03-01"), y = 8, label = "COVID"), 
            color = "Black", size = 4, angle = 0, fontface="italic") +  
  xlab("Date") +
  ylab("GDP Period Average") +
  theme_light()

p
```

```{r}
ggsave(filename = "GDP PA synthetic (UR & Employed & Business Income).png", plot = p, width = 8, height = 6, dpi = 300)
```

```{r}
plotting_ds <- plotting_ds %>%
  rename(
    "synthetic GDP quarterly growth rate (KS)" = "synthetic GDP",
    "synthetic GDP quarterly growth rate (KF)" = "synthetic GDP 2"
  )

```

```{r}
library(writexl)

# Export to Excel
write_xlsx(as.data.frame(smoothings), "C:.../PD_series/outputs/smoothed series.xlsx")
write_xlsx(series, "C:.../PD_series/outputs/input series.xlsx")
write_xlsx(plotting_ds, "C:.../PD_series/outputs/Synthetic GDPs (quarterly rates) UR, Employed, Business Income.xlsx")
```
